\section*{Appendix 4: Installing \pired on a Host System}
\addcontentsline{toc}{section}{Appendix 4: Installing \pired on your System}

The following primarily concerns the installation of \pired$^+$, i.e., of the
 compiled graph reducer version of \pired, which is our standard version for
release to the public. The installation of the interpreter
 version \pired$^*$ is very similar.

\pired$^+$ will be made available to you on a floppy disc or on our ftp-server as a compressed
 {\tt tar} file. In order to get it loaded into your system, you must
execute, under the directory under which you wish to install it, the
UNIX commands 
\begin{verbatim}
           tar  xvf  /dev/rfd0c 
           uncompress  red.tar.Z 
           tar  xvf  red.tar
\end{verbatim}
in this sequence. This will open up a new subdirectory {\tt red/} under
 which you can find all the (executable) files and further subdirectories 
necessary to compile and assemble the complete reduction system. 

If you have under the same directory another version of \pired configured for a
different target machine, first execute
\begin{verbatim}
            make  clean
\end{verbatim}
to delete all files that may be invalid. Also, on a UNIX system V/4 
make sure that {\tt /usr/bin} precedes  {\tt usr/ucb} in the directory path
since the {\tt cc} command in the latter uses libraries which have bugs.

If you wish to install \pired on either a SUNSPARC or an Apollo system,
you simply need to run next the executable file 
\begin{verbatim}
            configure  sun|apollo          .
\end{verbatim}
in order to prepare both the editor and the run-time system 
 for compilation to executable code.
For all other systems, you may have to set `by hand' several flags in two
{\tt makefiles} named {\tt fed/Makefile} and {\tt src/Makefile}. These
compiler flags are as follows:

\vspace{5mm}

\noindent Editor flags contained in {\tt fed/Makefile} (with default setting):
 
\vspace{4mm}

\begin{tabular}{ll}
{\tt -DUNIX=1} & for UN*X OS systems;\\
{\tt -DAPOLLO=0} & for Apollo systems set this to 1;\\
{\tt -DTRICK1\_AP=1} & for internal use only (don't change);\\
{\tt -DKEDIT=0} & for use with a distributed version of \pired (don't change);\\
{\tt -DM\_PATT=1} & for multiple pattern matches;\\
{\tt -DTRACE=0} & for internal use only (don't change);\\
{\tt -DClausR=0} & for extended system version with logical variables (don't 
change);\\
{\tt -DB\_DEBUG=0} & for internal use only (don't change);\\
\end{tabular}
\\
\newpage
\noindent Run-time flags contained in {\tt src/Makefile} (with default setting):

\vspace{5mm}

\begin{tabular}{ll}
{\tt -DUNIX=1} & for UN*X OS systems;\\
{\tt -DODDSEX=0} & for Littleendian systems set this to 1;\\
{\tt -DAPOLLO=0} & for Apollo systems set this to 1;\\
{\tt -DSYS5=0} & for System5 systems set this to 1;\\
{\tt -DPI\_RED\_PLUS=1} & for internal use only (don't change);\\
{\tt -DRED\_TO\_NF=1} & enables reductions to normal forms;\\
{\tt -DDEBUG=0} & disables custom-made debug system;\\
{\tt -DDBUG=0} & disables fred-fish's debug package;\\
\end{tabular}

\vspace{5mm}

\noindent Having executed the {\tt configure*} file or set the
 compiler flags directly, you may now call
\begin{verbatim}
            make
\end{verbatim}
which finally produces the executable file {\tt reduma*}. Calling it brings \pired 
into existence on your host system, i.e., the user interface 
shows up on the screen (window) you are
working with.

Before calling {\tt make}, it is advisable to look into the initialization file {\tt red.init} in order to make sure that the system  is
set up with the parameters as you  need them for most of your
 applications. This concerns primarily the sizes of the various
 run-time structures (heap size, stack sizes, etc.). This is good 
practice insofar as any changes of these parameters 
 by 
the respective system commands do not survive a
session, i.e., whenever you call the system with the command 
{\tt reduma}, it is initialized with the old parameters taken from
{\tt red.init} at compile time. 

The installation of the system version {\sc Xred} which features an 
X-window frontend pre-supposes the existence of the system \pired, preferably
under the same directory. It also requires the following system environment:
\begin{itemize}
\item a UNIX system V, release 3.2 with IPC message queues enabled;
\item an X-window system X11R5 or higher;
\item X-view 3.2 or higher;
\item a colour work station or X-terminal;
\end{itemize}

As with \pired, {\sc Xred} will be made available either on a floppy disc or on
 our ftp-server as a compressed {\tt tar} file. To get it installed, you must
execute, under this directory, the UNIX commands 
\begin{verbatim}
           tar  xvf  /dev/rfd0c 
           uncompress  udi.tar.Z 
           tar  xvf  udi.tar
\end{verbatim}
in this sequence, which creates a new subdirectory {\tt udi/} of source files. In order
to get everything compiled, first you have to climb down into this directory with
{\tt cd udi}, under which you will find a file {\tt README} (which should be read before proceeding)
and a file {\tt Install}. In {\tt Install} you will have to set a number of paths
according to your particular installation.
These are

\begin{tabular}{ll}
{\tt REDLIB} & = path to the \pired library;\\
{\tt REDSRC} & = path to the source code of \pired\\
{\tt SFEDSRC} & = path to the source code of the syntax-drected editor;\\
{\tt XFEDLIB} & = path to the X-window version of the library of the
\\ & syntax-directed editor;\\
{\tt TOPPATH} & = the complete path to the directory under which the
\\ & {\tt Install} file is
held;\\
{\tt SFED\_PATH} & = search path to the help file {\tt creduma.h} of the
syntax-\\ & directed editor.\\
\end{tabular}
 
Next you have to call this installation script with {\tt ./Install}. Upon termination,
it shows how the following two environment variables are set up:
\begin{tabular}{ll}
{\tt \$PATH} & which should contain the directory {\tt ./bin};\\
{\tt \$FHELPFILE} & which must point to the help file {\tt creduma.h}
for the syntax-\\ & directed editor.\\
\end{tabular}
If not, you must set it according to your particular installation.

Finally you may compile your source files using the following sequence of commands:
\begin{verbatim}
     cd src
     make depend
     make
\end{verbatim}

\noindent This will create in your directory
\begin{itemize}
\item  {\tt udi/bin} the executable files
\begin{verbatim}
        Xred
        kir
        marco
        sxred
\end{verbatim}
\item {\tt udi/lib} the libraries and example make files
\begin{verbatim}
        Example.Makefile.Compiler-Edoitor
        Example.Makefile.CompilerGen
        editor.o
        libkir.o
        libmarco.o
        marco-main.o
\end{verbatim}
\end{itemize}
\noindent A brief explanation of these files may be found in the {\tt README} file.  
