
This README describes the installation and compilation of :

 - marco (Make Another Reduction COmpiler)
   marco is a 2 pass compiler generator for functional languages which use
   the pi-red reductionsystem for program execution. A marco generated compiler
   translates high-level functional program into low-level Orel/0 programs.

   The marco source (./src/marco) is splitted into three sections :
   
    * First the compilergenerator (compgen) which translates a compiler-
      specification in a compiler.

    * Second the compilerlibrary (library) which is a compilerindependent function-
      library.

    * Third : Two examples (examples) compilerspecification for macro.

      + kir  : A macro-generated KiR (KIel Reduction Language).
      + calc : A simple calculator tool.

 - Xred (This XView frontend for the syntax-directed editor)
   The source for this X-windows-based frontend is in ./src/xv

 - Some help functions (sources) :

   * clt_srv : A client-server module for the syntax-directed editor and the
     X-Windows frontend.

   * dbug : Fred Fish DBUG package for internal debugging purposes.

and two other program packages, which are NOT part of this archive :

 - fed
   The syntax directed editor for the pi-red reduction system.

 - red
   The pi-red reductionssystem.

Before you can compile a program from this packages, you have to 
install first the fed and red packages !!

----------------------------------------------------------------------------------
REQUIRMENTS :
----------------------------------------------------------------------------------

This packages requires following stuff :

 - Unix System V 3.2
 - X11 R5 system
 - XView 3.2 Includes & Library in your library path
 - System V IPC message queues

To run the Xred program your MUST have the server version of the syntax-directed
editor named "sxred" in your binary-search path.

Without these requirments the programs will NOT run correct.

----------------------------------------------------------------------------------
INSTALLATION :
----------------------------------------------------------------------------------

Setup your pathes in the Install file and run this script. To compile
the sources you must change your directory to src. Then type 

 make depend
 make

After compilation are in the toplevel path bin links to the generated programs.
To use this programs you have to copy this files to your binary search path.

----------------------------------------------------------------------------------
PROGRAMS : 
----------------------------------------------------------------------------------

After succesfull compilation you will get 4 programs in ./bin :

 * Xred    : X-window frontend for sxred.
 * sxred   : Serverversion of fed.
 * kir     : A KiR compiler.
 * marco   : Tool for building the KiR-compiler.

Copy these programs to your binary search path.

----------------------------------------------------------------------------------
DEINSTALLATION :
----------------------------------------------------------------------------------

To cleanup run the DeInstall script.

-- 
Ulrich Diedrichsen      udi@informatik.uni-kiel.d400.de
Oooooh, eight carousels!  We're in for a real treat!
 <from> Marge welcomes her sisters, laden with slides of their latest trip,
        ``Krusty Gets Busted''

